---
title: "Getting Cell Tower Locations"
author: "[Truc Viet 'Joe' Le](mailto:tjle@andrew.cmu.edu)"
date: "March 26, 2015"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(ig.width=6, fig.height=6, fig.path='./figures/mobile/',
                      warning=FALSE)
```

This tutorial demonstrates how to match `cell_id`'s in the **mobile phone** dataset with their corresponding **real-world** spatial locations (i.e., longitude and latitude coordinates). Note that not all `cell_id`'s can be matched and located. Ignore those that are unmatched, which turns out to be **a lot**.

First, make sure that you have successfully cloned the [GitHub repository](https://github.com/vietexob/mobile-intelligence) that hosts the project. If you're new to **git**, look for brief [instructions here](https://help.github.com/articles/fork-a-repo/). Then, go to the cloned directory, it should be named "mobile-intelligence" by default. This is **important**: Use **RStudio** to open the project file `mobile-intelligence.Rproj` in the directory. You should have now set the path correctly. Now, load the useful packages as always. We will use them for retrieving the data and plotting them on a map.

```{r load_packages, include=TRUE}
library(rmongodb)
library(ggplot2)
library(ggmap) ## for plotting maps
```

Note that if a package is not installed, you will get an error when trying to load it. You need to install that package before you can load it. For example:

```{r install_packages, eval=FALSE}
install.packages("ggmap", dependencies=TRUE)
```

The spatial locations of the cell towers are stored in a separate CSV file. Therefore, we'll need to load it and read the cell names such that they can be matched to those `cell_id`'s in the database. The way to match those is to read only the last 4 letters of each cell name and convert it to the corresponding decimal representation (from hexadecimal).

```{r read_locations}
## Read the cell towers' spatial coordinates from the CSV file
cell.towers <- read.csv(file="./data/mobile/cell_coord.csv", stringsAsFactors=FALSE)
cell.names <- cell.towers$Cell # retrieve the cell names
## Take only the last 4 letters for the cell names
cell.names.short <- vector()
for(i in 1:length(cell.names)) {
  cell.names.short[i] <- substr(cell.names[i], start=4, stop=nchar(cell.names[i]))
}
## Convert from hexadecimal into decimal format
cell.names.dec <- strtoi(cell.names.short, 16L)
```

We now connect to the remote server as we did before.

```{r server_connection}
## Login credentials
host <- "heinz-tjle.heinz.cmu.edu"
username <- "student"
password <- "helloWorld"
db <- "admin"
## Connect to MongoDB remote server
mongo <- mongo.create(host = host, db = db, username = username, password = password)
## Check if we are successfully connected
mongo.is.connected(mongo)
## The database we're working with is 'admin' and the collection is 'cellular'
collection <- "cellular"
namespace <- paste(db, collection, sep=".")
```

We retrieve **all** the `cell_id`'s and convert them into hexadecimal format.

```{r get_cell_id}
## Get the cell tower locations
cell.loc <- mongo.distinct(mongo, namespace, "cell_id")
## Convert into vector
cell.loc <- unlist(cell.loc)
## Convert from hexadecimal into decimal
cell.loc.dec <- strtoi(cell.loc, 16L)
```

We will now match each `cell.loc.dec` (i.e., the decimal representation of a `cell_id`) to a corresponding element in the `cell.names.dec` vector: they are matched if they have the same decimal number representation. We will also create a mapping (i.e., a [hash table](http://en.wikipedia.org/wiki/Hash_table)) called `cell_id.coord.rowIndex` that maps each `cell_id` in its original hexadecimal format to a corresponding row index of the `cell.towers` data frame. This will come in handy when later we wish to look up for a `cell_id`'s longitude and latitude coordinates. We will also create a data frame called `cell.loc.data` in order to visualize those locations on a map (more on that later). Finally, we also want to keep track of how many locations in our database that can be matched with the coordinates given in the CSV file. We will see that only a **very small** fraction can be matched.

```{r cell_matching}
## Create a mapping from each cell_id to the corresponding cell.towers table's row index
cell_id.coord.rowIndex <- new.env() # this is how a hash table is declared in R
## Create a data frame to store and visualize the locations
cell.loc.data <- data.frame(stringsAsFactors=FALSE)
counter <- 0 # this counter keeps track of the number of matches
for(i in 1:length(cell.loc)) { # iterate through each retrieved cell location
  if(cell.loc.dec[i] %in% cell.names.dec) { # check if there is a match
    rowIndices <- which(cell.names.dec == cell.loc.dec[i]) # find the matched index
    if(length(rowIndices) == 1) { # if there is one unique match
      rowIndex <- rowIndices
    }
    else { # otherwise, matched with more than one row
      ## Take the first one that's matched
      rowIndex <- rowIndices[1]
    }
    
    ## Get the long and lat coordinates
    longitude <- as.numeric(cell.towers$Longitude[rowIndex])
    latitude <- as.numeric(cell.towers$Latitude[rowIndex])
    if(!is.na(longitude) && !is.na(latitude)) { # some of the coords are corrupted
      cell_id.coord.rowIndex[[toString(cell.loc[i])]] <- rowIndex
      loc.row.data <- data.frame(cell=cell.loc[i], longitude=longitude, latitude=latitude)
      cell.loc.data <- rbind(cell.loc.data, loc.row.data)
      counter <- counter + 1
    }
  }
}
## How much percent is matched?
(round(counter / length(cell.loc) * 100, 2))
```

Our result shows that only 4.37% of the `cell_id`'s in our database can be matched to the real-world coordinates. That is about 2,252 unique locations, which is quite good enough to put them all on a map. But before that, I want to demonstrate how the mapping `cell_id.coord.rowIndex` can be very useful when we know a `cell_id` and we wish to quickly retrieve its coordinates.

```{r get_coords}
cell_id <- "1451" # suppose we know this cell_id
rowIndex <- cell_id.coord.rowIndex[[cell_id]]
(longitude <- cell.towers$Longitude[rowIndex])
(latitude <- cell.towers$Latitude[rowIndex])
```

We are now ready to map those 2,252 locations in the `cell.loc.data` data frame using their longitude and latitude coordinates. The following code will visualize them on a static map retrieved from Google Maps.

```{r get_map}
## Retrieve a map from Google Maps with center at the means of all the coordinates
cell.loc.map <- get_map(location = c(lon = mean(cell.loc.data$longitude),
                                     lat = mean(cell.loc.data$latitude)),
                        zoom = 10, scale = 2) # scale specifies the resolution of the map
## Make a map that plots each cell tower using a little red dot
cell.tower.pts <- geom_point(data = cell.loc.data, aes(x = longitude, y = latitude,
                                                       fill = "red", alpha = 0.80),
                             size = 1, shape = 21)
## This line overlays the cell tower locations onto the retrieved map
cell.loc.map <- ggmap(cell.loc.map) + cell.tower.pts + guides(fill = FALSE, alpha = FALSE,
                                                              size = FALSE)
## Give the map a title
cell.loc.map <- cell.loc.map + ggtitle("Locations of Cell Towers")
## Add the density contours
cell.loc.map <- cell.loc.map + geom_density2d(data = cell.loc.data,
                                              aes(x = longitude, y = latitude))
## Print the map
print(cell.loc.map)
```

Finally, if you want to save the map (image) onto disk for use later, the following code will do the job.

```{r save_map}
## Save the plot on disk, specify dimension 8x8 inches
ggsave(filename="./figures/mobile/cell_towers.png", width=8, height=8)
```

This is the end of the tutorial. E-mail me should you run into any problems.
